A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-04-19 paper/paper1/ms.tex
Add text about sub Saturn runaway to the intro

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may effectively migrate to close-in orbital separations, but once they arrive, the intense irradiation overheats the planet.  This heating leads to runaway inflation and ultimately complete disintegration.  In this scenario, the inflationary half-life may be so short as to lead to a low probability of observing members in the class, and hence the apparent lack of inflated sub-Saturns.

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Here, lightly ionized atmospheric flows induce drag in a planetary magnetic field, weakly coupling the stellar incident energy into the planetary interior.  \citet{2018AJ....155..214T} recognized the above two explanations for the lack of sub-Saturns, while statistically favoring Ohmic dissipation as the mechanism responsible for inflating Hot Jupiters.  Therefore Ohmic dissipation stands out as a leading physical mechanism unifying several observations.  Ohmic dissipation does not preclude the existence and role of other heating mechanisms---such as photoevaporation and tidal heating---but it does stand out as having made successful predictions before data were widely available to test them.

A runaway inflation scenario implies an inevitable and profound mass loss rate for inflated hot Saturns.  The planet's atmosphere overflows its Roche lobe at an ever-increasing pace, yielding instantaneous mass loss rates possibly much larger than we have seen in previous long-term stable systems.  The ability to detect that conceivable mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own limitations.  Ly $\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages over Ly$\alpha$.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

@@

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU and a 4.8-day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  It orbits a possibly evolved subgiant, opening the prospect for ``re-inflation'' \citep{2022AJ....163...53S,2022AJ....163..120G,2023arXiv230306728G} as the planet's insolation increased with the star's rapid luminosity increase.  The tidal gravity of the nearby massive host star may accelerate mass loss rates even further \citep{2007A&A...472..329E}. Its status as a rare inflated Hot Saturn makes \object{HAT-P-67 b} a promising laboratory, including but not limited to, testing the inflationary predictions of the Ohmic dissipation mechanism.

@@

Excess is confidently detected at from phases $-$0.3 to $+$0.03.  The full-width at half maximum (FWHM) of the feature is about 1.8-2.6~\AA, equivalent to a line-of-sight velocity distribution in the range of 50-75~km/s.  Individual fits to the He feature show a typical line center uncertainty $\pm0.1$~\AA~ or better.  The out-of-transit phases exhibit a line center position of 10833.4~\AA, consistent with the stellar \ion{He}{1} rest frame zero velocity.  The excess absorption resides slightly blueward of the stellar zero restframe velocity reference.  As mentioned previously, the telluric masking near 10836 \AA~ censors our view of the existence (or not) of any redshifted lobe in this region.  A more careful treatment of the telluric absorption could rescue some of the data in this region.  The excess detections exhibit a bulk blueshift of up to 30~km/s.

>>>>

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may either form \emph{in-situ} or effectively migrate to close-in orbital separations \citep{2018ARA&A..56..175D}, but once they arrive, the intense irradiation overheats the planet.  This heating leads to runaway inflation and ultimately complete disintegration.  In this scenario, the inflationary half-life may be so short as to lead to a low probability of observing members in the class among the overall population, and hence the apparent lack of inflated sub-Saturns.

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Here, lightly ionized atmospheric flows induce drag in a planetary magnetic field, weakly coupling the stellar incident energy into the planetary interior.  \citet{2018AJ....155..214T} recognized the above two explanations for the lack of sub-Saturns, while statistically favoring Ohmic dissipation as the mechanism responsible for inflating Hot Jupiters.  Therefore Ohmic dissipation stands out as a leading physical mechanism for inflation and mass loss.

Recently, \citet{2023ApJ...945L..36T} showed that hot Saturns can undergo catastrophic erosion by stellar X-ray and Extreme UV (XUV) photoevaporative mass loss. Planets with densities less than $\sim$1 g~cm$^{-3}$ achieve mass loss rates of up to $10^3$ $M_\oplus /$Gyr setting up a positive feedback loop: the planet increases in radius, overflowing its Roche lobe, fueling greater mass loss, and increasing in radius even further.  This vicious cycle systematically depopulates the mass-radius plane with a cliff defined by the $\sim$1 g~cm$^{-3}$ dividing line.

A runaway inflation scenario predicts an inevitable and profound mass loss rate for inflated hot Saturns.  As the planet's atmosphere overflows its Roche lobe at an ever-increasing pace instantaneous mass loss rates may exceed those previously seen in atmospheric escape measurements to date.

The ability to detect that conceivable mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own limitations.  Ly $\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

@@

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU and a 4.8-day orbital period \citep{2017AJ....153..211Z}.  Its strong insolation ($T_\mathrm{eq}\sim2000\;$K) combined with \object{HAT-P-67 b}'s extremely low surface gravity makes it an exceptional candidate for strong atmospheric mass loss via Roche lobe overflow.  Its $\sim$0.05 g~cm$^-3$ density places it below the 1.0 g~cm$^-3$ threshold predicted to exhibit runaway inflation by \citet{2023ApJ...945L..36T}.  It orbits a possibly evolved subgiant, opening the prospect for ``re-inflation'' \citep{2022AJ....163...53S,2022AJ....163..120G,2023arXiv230306728G} as the planet's insolation increased with the star's rapid luminosity increase.  The tidal gravity of the nearby massive host star may accelerate mass loss rates even further \citep{2007A&A...472..329E,2023ApJ...945L..36T}. Its status as a rare inflated Hot Saturn makes \object{HAT-P-67 b} a promising laboratory, including testing the runaway inflationary predictions of Ohmic dissipation and XUV photoevaporation.

@@

Excess is confidently detected from phases $-$0.3 to $+$0.03.  The full-width at half maximum (FWHM) of the feature is about 1.8-2.6~\AA, equivalent to a line-of-sight velocity distribution in the range of 50-75~km/s.  Individual fits to the He feature show a typical line center uncertainty $\pm0.1$~\AA~ or better.  The out-of-transit phases exhibit a line center position of 10833.4~\AA, consistent with the stellar \ion{He}{1} rest frame zero velocity.  The excess absorption resides slightly blueward of the stellar zero restframe velocity reference.  As mentioned previously, the telluric masking near 10836 \AA~ censors our view of the existence (or not) of any redshifted lobe in this region.  A more careful treatment of the telluric absorption could rescue some of the data in this region.  The excess detections exhibit a bulk blueshift of up to 30~km/s.
